---
title: "Benchmarking cloud detectors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking cloud detectors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cloudscape)
set.seed(3)
```

## Why simulate at all

Hand-labelled benchmarks have a ceiling. Annotators disagree at thin-cloud
boundaries, shadow is the weakest-labelled class in every published dataset,
and the scenes chosen for labelling are rarely the surfaces where masks
actually fail. Simulation is complementary, not a replacement: truth is exact,
and difficulty is a dial rather than a property of whichever scenes someone
labelled.

## A controlled scene

```{r}
bg <- list(blue = matrix(0.06, 128, 128), green = matrix(0.08, 128, 128),
           red = matrix(0.05, 128, 128), nir = matrix(0.32, 128, 128),
           swir16 = matrix(0.20, 128, 128), swir22 = matrix(0.12, 128, 128))

sim <- cl_simulate(background = bg, coverage = 0.28, opacity = c(0.35, 1),
                   height = 2200, res = 30, sun_zenith = 42,
                   sun_azimuth = 155, seed = 7)
c(realised_coverage = round(mean(sim$cloud), 3),
  shadow_pixels = sum(sim$shadow))
```

Opacity is the difficulty control. Fully opaque cloud is easy for every method;
methods separate between roughly 0.2 and 0.6, which is where thin cirrus and
cloud edges live and where published masks disagree most.

## Comparing methods, with uncertainty

```{r}
m_thr <- cl_probability(sim$bands, "threshold", sensor = "sentinel-2-msi")
m_fm  <- cl_probability(sim$bands, "fmask",     sensor = "sentinel-2-msi")

cmp <- cl_compare(list(threshold = m_thr$class, fmask = m_fm$class),
                  sim$cloud, n_boot = 300)
cmp$summary
cmp$pairwise[, c("method_a", "method_b", "mcnemar_chi2", "p_value")]
```

A ranking without an interval is not a result. Two methods differing by a few
tenths of a percent on one scene are not distinguishable, and McNemar's test on
the paired outcomes is the right instrument because the same pixels are scored
by both.

For spatially autocorrelated masks a pixel-wise bootstrap understates
uncertainty. Pass `block =` with tile identifiers for an honest interval.

## Discrimination is not calibration

```{r}
c(auc   = round(attr(cl_roc(m_fm$probability, sim$cloud), "auc"), 3),
  brier = round(attr(cl_calibration(m_fm$probability, sim$cloud), "brier"), 4),
  ece   = round(attr(cl_calibration(m_fm$probability, sim$cloud), "ece"), 4))
```

A method can rank pixels perfectly and still emit probabilities that mean
nothing, in which case they cannot be used to weight observations or propagate
uncertainty. Reporting AUC alone hides this. During development of this
package, an early Fmask implementation scored AUC 1.000 with expected
calibration error 0.373 — the ranking was perfect and the threshold was in
entirely the wrong place. The calibration diagnostic caught it; accuracy at the
default threshold was 0.61.

## Disagreement without labels

```{r}
cl_disagreement(list(threshold = m_thr$class, fmask = m_fm$class))$pairwise
```

This needs no reference labels at all, which is what makes it deployable
globally. Passing `strata =` — a land-cover or snow layer — reports where
methods diverge, and those are the regions where a derived cloud statistic
should carry a caveat.

## A difficulty gradient

```{r}
bench <- do.call(rbind, lapply(c(0.15, 0.3, 0.5, 0.8, 1.0), function(op) {
  s <- cl_simulate(background = bg, coverage = 0.3, opacity = op, seed = 21)
  m <- cl_probability(s$bands, "fmask", sensor = "sentinel-2-msi")
  v <- cl_validate(m$class, s$cloud)
  i <- which(v$class == "1")
  data.frame(opacity = op, recall = round(v$recall[i], 3),
             precision = round(v$precision[i], 3), iou = round(v$iou[i], 3))
}))
bench
```

Reporting a single accuracy number for a cloud detector conceals exactly this
curve. Two methods with identical overall accuracy can behave completely
differently on thin cloud, which is where most of the disagreement in
operational products lives.

## Using real benchmarks

```{r, eval = FALSE}
cl_benchmark("cloudsen12", n = 50)
```

Benchmark imagery ships in the companion data package. `cl_benchmark("synthetic")`
generates scenes on the fly and needs nothing.
