---
title: "Getting started with cloudscape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cloudscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cloudscape)
```

## The question this package is for

Most cloud software answers "which pixels are cloudy". That is a means, not an
end. The question behind it is usually "can I do my study here", and answering
that needs three things the mask alone does not give you: an unbiased spatial
unit, an honest account of what scene metadata can and cannot tell you, and a
translation from cloud statistics into the units your conclusions are stated
in.

## An unbiased spatial unit

Scene footprints are the wrong unit for statistics. A WRS-2 scene covers less
ground area at high latitude, and neighbouring paths overlap more, so counting
images per path/row conflates acquisition density with geometry. It also makes
Landsat and Sentinel-2 incomparable, since their footprints differ.

```{r}
g <- cl_grid(res = 100000)
box <- function(lat) cbind(c(0, 2, 2, 0, 0), c(lat, lat, lat + 2, lat + 2, lat))
nrow(cl_grid_index(box(50), g))   # cells covered at 50N
nrow(cl_grid_index(box(70), g))   # same 2x2 degrees at 70N
```

Every cell has the same true ground area, so counts are directly comparable
across latitude and across sensors.

## Units are enforced, not assumed

```{r, error = TRUE}
cl_stats(1:2, "cloud_fraction", c(0.3, 1.4))
```

```{r}
s <- cl_stats(1:3, "n_scenes", c(20, 21, 22), period = "2023",
              denominator = "cell-year")
s$denominator
```

The denominator is part of the data. A total across ten years divided by ten is
a mean *per year*; if it is labelled as a mean per scene, the numbers become
physically impossible for the stated unit, and nothing in a plain data frame
would catch that.

## From cloud fraction to usable observations

Scene-level cloud cover is a scene average. Turning it into a per-pixel clear
probability requires an assumption, and the naive one is optimistic:

```{r}
d <- seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "5 days")
ts <- cl_simulate_series(d, p_cloud = 0.55, persistence = 0.35, seed = 2)
obs <- cl_obs(cell = 1, date = ts$date, cloud_fraction = ts$cloud_fraction,
              sensor = "sentinel-2-msi")
suppressMessages(head(cl_clear_prob(obs, "independent"), 3))
head(cl_clear_prob(obs, "clustered"), 3)
```

Cloud is clustered, so a scene reported as half cloudy is more often half
covered by one deck than uniformly obscured. Tier-A results are upper bounds,
and the package says so rather than leaving it implicit.

## What actually breaks a time series

Not the mean. The longest stretch without a usable image.

```{r}
cl_stats_wide(cl_gaps(obs, threshold = 0.2, critical = 30))
```

## Translating into days

```{r}
res <- lapply(c(0.3, 0.7, 0.9), function(p)
  cl_pheno_power(d, cloud_fraction = p, n_sim = 200, seed = 4)$summary)
do.call(rbind, res)[, c("n_usable_median", "sos_mae", "failure_rate")]
```

The failure rate is often the more important column. In heavily clouded regions
a study does not return a biased green-up date, it returns none at all.

## Constellations change through time

```{r}
sy <- cl_synergy(c("landsat-8-9-oli", "sentinel-2-msi"),
                 "2014-01-01", "2025-12-31")
sy[sy$sensor == "COMBINED" & sy$period %in% c("2014", "2018", "2022", "2025"),
   c("period", "n_platforms", "effective_revisit_days")]
```

Sampling density is not stationary. Landsat 9 joined in late 2021 and
Sentinel-2C in 2024, so a trend computed across such a period partly reflects
acquisition capacity rather than the environment. Any long time-series analysis
should account for this explicitly.

## Where to go next

* `vignette("detection")` — cloud and shadow detection, and comparing methods
* `vignette("benchmarking")` — simulation-based evaluation
* `vignette("phenology")` — designing a phenology study around cloud
