---
title: "Cloud and shadow detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cloud and shadow detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cloudscape)
set.seed(5)
```

## Methods are registered, not hard-coded

```{r}
cl_methods()
```

Each method declares the bands it needs, so applicability can be checked before
anything is downloaded:

```{r}
cl_methods(sensor = "sentinel-2-msi")$method
cl_methods(sensor = "generic-multispectral")$method
```

A four-band UAV camera has no SWIR, so Fmask-style detection is unavailable and
the package says so rather than silently producing nonsense.

## Adding your own method

```{r}
cl_method_register(
  "blue_ratio",
  requires = c("blue", "red"),
  type = "statistical",
  reference = "example",
  fun = function(bands, geom, drv, cut = 1.3, ...) {
    r <- bands$blue / pmax(bands$red, 1e-6)
    1 / (1 + exp(-(r - cut) / 0.1))
  })

"blue_ratio" %in% cl_methods()$method
```

Nothing else changes. `cl_probability()`, `cl_compare()` and the evaluation
tools work with it immediately, because they address methods through the
registry rather than by name.

## Reading the producer's own mask

```{r}
qa <- matrix(0L, 4, 4)
qa[1, ] <- bitwShiftL(1L, 3)   # cloud bit
qa[2, ] <- bitwShiftL(1L, 4)   # cloud-shadow bit
dec <- cl_qa_decode(qa, "landsat-8-9-oli")
dec$class
```

The same call works for Sentinel-2, whose SCL is categorical rather than
bit-packed, because the decoding rule lives in the sensor driver. Both are
translated into the one vocabulary given by `cl_classes()`, which is what makes
CFMask, Sen2Cor and a neural mask directly comparable.

## Shadow geometry

```{r}
cl_shadow_offset(height = 2000, sun_zenith = 45, sun_azimuth = 180)
cl_shadow_offset(height = 2000, sun_zenith = 45, sun_azimuth = 90)
```

Sun in the south puts the shadow to the north; sun in the east puts it to the
west. Displacement is `height * tan(zenith)`, with an optional correction for
the satellite not viewing from nadir — negligible for Landsat, worth applying
across a Sentinel-2 swath.

When a thermal band exists, cloud height can be estimated directly:

```{r}
bt <- matrix(300, 40, 40)
cloud <- matrix(FALSE, 40, 40); cloud[15:25, 15:25] <- TRUE
bt[cloud] <- 282
cl_cloud_height(bt, cloud)
```

Without a thermal band the honest answer is a search range, not a number, and
that is what is returned.

## Terrain shadow is not cloud shadow

```{r}
dem <- matrix(0, 40, 40); dem[, 30:40] <- 600
sum(cl_terrain_shadow(dem, sun_zenith = 75, sun_azimuth = 90, res = 30))
sum(cl_terrain_shadow(dem, sun_zenith = 75, sun_azimuth = 270, res = 30))
```

In rugged terrain both are dark and both correlate with aspect. Neither Fmask,
Sen2Cor's SCL, nor single-date neural masks separate them explicitly, so
topographic shadow is routinely absorbed into cloud-shadow statistics. Computing
it from a DEM lets `cl_shadow()` exclude it.

## Where masks should not be trusted

```{r}
bands <- list(blue = matrix(0.2, 20, 20), green = matrix(0.35, 20, 20),
              red = matrix(0.3, 20, 20), nir = matrix(0.4, 20, 20),
              swir16 = matrix(0.05, 20, 20))
r <- cl_reliability(bands)
names(r$flags)
round(mean(r$reliability), 3)
```

High NDSI with low SWIR is snow, which every published cloud mask confuses with
cloud to some degree. Flagging it means a derived statistic can carry an honest
caveat instead of an unqualified number.
