---
title: "Designing a phenology study around cloud"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing a phenology study around cloud}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(cloudscape)
set.seed(1)
```

## The question

You want land-surface phenology from optical imagery. Before committing, you
want to know whether the site will yield a defensible green-up date, and if
not, what would change that.

Mean cloud cover cannot answer this. Two sites with identical 60% annual cloud
can differ by a factor of three in retrieval error, depending on *when* the
cloud falls and how it clusters.

## Step 1: how many observations, and when

```{r}
dates <- seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "5 days")
site <- cl_simulate_series(dates, p_cloud = 0.62, persistence = 0.4, seed = 11)
obs <- cl_obs(cell = 1, date = site$date, cloud_fraction = site$cloud_fraction,
              sensor = "sentinel-2-msi")

cl_stats_wide(cl_gaps(obs, threshold = 0.2, critical = 21))
```

The maximum gap matters more than the count. A 40-day gap that happens to
straddle green-up removes the transition entirely, and no amount of data later
in the season recovers it.

## Step 2: translate into days of error

```{r}
p <- cl_pheno_power(dates, cloud_fraction = site$cloud_fraction,
                    n_sim = 300, seed = 2)
p$summary[, c("n_usable_median", "sos_mae", "sos_iqr", "sos_bias",
              "failure_rate")]
```

Read the failure rate first. It is the fraction of years in which the
double-logistic fit does not converge at all. A study that fails to produce an
estimate is a different problem from one that produces a biased estimate, and
the two get confused when only mean error is reported.

## Step 3: how sensitive is this to clustering?

```{r}
grid <- expand.grid(cloud = c(0.4, 0.6, 0.8), rho = c(0, 0.35, 0.7))
res <- do.call(rbind, Map(function(cl, rho) {
  s <- cl_pheno_power(dates, cloud_fraction = cl, persistence = rho,
                      n_sim = 250, seed = 5)$summary
  data.frame(cloud = cl, rho = rho, usable = s$n_usable_median,
             sos_mae = s$sos_mae, failure = round(s$failure_rate, 3))
}, grid$cloud, grid$rho))
res
```

Along any row of constant cloud, the number of usable observations barely
moves, but the failure rate climbs with `rho`. Clustering does not remove
observations; it removes them from where they matter. Any feasibility
assessment that assumes independent cloud loss will be optimistic, and the more
so the more clustered the local climate.

## Step 4: would adding a sensor fix it?

```{r}
sy <- cl_synergy(c("landsat-8-9-oli", "sentinel-2-msi"),
                 "2015-01-01", "2025-12-31")
sy[sy$period == "2025", c("sensor", "n_platforms", "combined_obs",
                          "effective_revisit_days")]
```

Adding Landsat to Sentinel-2 shortens effective revisit substantially. Whether
that converts into usable observations depends on whether the two constellations
are clouded at the same times, which they largely are — so the gain is real but
smaller than the revisit numbers alone suggest. Verify with the actual archive
rather than assuming.

## Step 5: state the design explicitly

```{r}
cl_feasibility(p_clear = 0.38, n_acquisitions = 73, n_required = 15,
               window_days = 365, max_gap = 30)
```

## What to report in a paper

State the acquisition count, the maximum gap, and the simulated retrieval error
with its failure rate, for each site or stratum. Reporting mean cloud cover
alone leaves reviewers unable to judge whether the phenological conclusions are
supportable, and leaves you unable to explain why one site behaved differently
from another.

A caveat worth carrying: everything above uses tier-A cloud fractions, which are
scene averages. They give an upper bound on usable observations. Recomputing
with `tier = "qa"` on the actual quality layers will lower these numbers, and
the gap between the two is itself informative.
